---
title: "Assignment 9"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Prepared by: 

  *   Josh Levine (jl2108)
  *   Harsh Patel (hkp49)
  *   Jaini Patel (jp1891) 
  *   Yifan Liao (yl1463) 
  *   Aayush Shah (avs93)

# False Discovery Rate (FDR) #

The FDR is known as the expected proportions of type I errors. A type I error is where we incorrectly reject the null hypothesis. FDR is closely related to FWER (family-wise error rate), the probability of making any type I error at all.

### FDR Formula: ###

FDR = E (V/R | R>0) P(R>0)

  where:
  
    *   V = Number of Type I errors
    *   R = Number of rejected hypotheses


Below is the function for FDR implementation:
```{r fdr function}
false_discovery_rate <- function(p_values, Q){
  
  sorted_p_values = sort(p_values) #Sorting the p-values -> Step1

  len = length(sorted_p_values)      #Number of p-values

  #Hypothesis 1

  hypothesis_1 = Q*c(1:len)/len

  #Hypothesis 2 - If not independent

  hep = len * sum(1/c(1:len))

  hypothesis_2 = (Q * c(1:len))/hep

  #the p-vlaues less than the hypothesis line are considered to be interesting

  is_interesting_true = (sorted_p_values < hypothesis_1)
   
  #compute the index below which the sorted p-values are less than the line

  threshold_index = max(which(is_interesting_true == "TRUE"))

  print(paste0("The Threshold Index: ", threshold_index)) #The no. of TRUE values

  p_star = sorted_p_values[threshold_index]

  
  hypothesis = c(1:len)

  plot(hypothesis, sorted_p_values, col="black")

  lines(c(1:threshold_index), sorted_p_values[c(1:threshold_index)],
        col = "red", type="o")

  lines(hypothesis, hypothesis_1, type="l", col="black")

  
  #list index of hypothesis which are interesting in the original unsorted list of p values

  temp = (p_values < p_star)

  fd = which(temp == "TRUE")

  print("Index of intersting p values in the unsorted list:")

  print(fd)

  

  #false rejection rate

  frr = -log(p_star)/length(fd)

  print(paste0("False Rejection Rate: ", frr))

  

  #################################################################################

  # NOTE: FRR depends on the Q value given as an input to the function

  # Here, the Q value was 0.05 hence the frr is approximately 0.05

  #################################################################################

}
```


Below is the code to call the fdr function:
```{r }
# Test Vector of p-values

vec <- c(1e-5*runif(100),runif(900))

Q = 0.05

false_discovery_rate(vec, Q)
```
