---
title: "Feed Foward Neural network"
output: html_document
---

## R Markdown

---

#Feed Foward Neural network

Researching deeper and deeper in machine learning fields, we have reached some new neural network models in the forecast package called nnetar. A single hidden layer neural network is the most simple neural networks form. In this kind of single hidden layer form, there is only one layer of input nodes that send weighted inputs to the next layer of receiving nodes. The nnetar function inside the forecast package fits a single hidden layer neural network model to a timeSeries.  The approach of this function model is to use lagged values of the time series as input data, reaching to a non-linear autoregressive model.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
if(!require(quantmod)) install.packages("quantmod")
if(!require(forecast)) install.packages("forecast")
if(!require(xlsx)) install.packages("xlsx")
if(!require(tseries)) install.packages("tseries")
if(!require(timeSeries)) install.packages("timeSeries")
if(!require(dplyr)) install.packages("dplyr")
if(!require(fGarch)) install.packages("fGarch")
if(!require(prophet)) install.packages("prophet")
if(!require(ruggarch)) install.packages("rugarch")
install.packages("PerformanceAnalytics")
install.packages("FinTS")
install.packages("strucchage")
install.packages("TSA")
suppressPackageStartupMessages(library(PerformanceAnalytics))
suppressPackageStartupMessages(library(FinTS))
suppressPackageStartupMessages(library(TSA))
suppressMessages(library(nnet))
library(prophet)
library(quantmod)
library(forecast)
library("xlsx")
library(tseries)
library(timeSeries)
library(dplyr)
library(fGarch)
library(rugarch)
library(ggplot2)
library(urca)
library(fBasics)
```

Load the data

```{r cars}
#Getting data
suppressMessages(getSymbols("^DJI", from = "2005-01-01", to = "2020-12-05"))
head(DJI)
tail(DJI)
```

We can apply Feed Foward Neural network models to the index or any stock price of our choosing by n(h)=n(s)/(αn(i)+αn(0)), where n(i) stands for number of input neurons; n(0) stands for number of output neurons; n(s) stands for number of train samples and α=1.5^(−10).

```{r}
dji_close <- DJI[,"DJI.Adjusted"]#closed value for dji
#With the hidden layers approach explained we proceed to calculate them:
alpha <- 1.5^(-10)
nh <- length(dji_close)/(alpha*(length(dji_close)+30))
Adjunclose <- DJI[,"DJI.Close"]
AdjunL1 <- Lag(Adjunclose,1)
#print(length(dji_close))
#Now with the hidden layers calculated we proceed to apply the nnetar function with the parameters selected.
lambda <- BoxCox.lambda(dji_close)
dnn_pred <- nnetar(dji_close, size= nh, lambda = lambda,MaxNWts=1200)#Fitting nnetar
neural.price.model<- nnetar(AdjunL1, lambda = lambda)
#We use a Box Cox lambda to ensure the residuals will be roughly homoscedastic. We forecast the next 30 days with the neural net fitted.
dnn_forecast <- forecast(dnn_pred, h= 30, PI = TRUE)
plot(dnn_forecast)#Fitting nnetar
```

#Artificial Neural Network

Then we continue working on Feed Foward Neural network. this time we found a multilayer perception which contains a three-layer with one hidden layer model trained with back-propagation algorithm. According to some research it can be the most effective to forecast financial time series.

```{R}
CloseL1 <- Lag(dji_close,1)# Preparing input data for ANN
#print(CloseL1[4000:4010])
#CloseL1 <- na.omit(CloseL1)
length <- length(dji_close)
CloseL1.train <- CloseL1[1:(length-30)]# Delimit training range
CloseL1.test <- CloseL1[(length-29):length]# Delimit testing range
```

The training parameters were set as follows: decay rate = 0.00001, number of units in the hidden layer = 10, and epoch size = 10000. 

```{R}
#N Net Function for Close Price
set.seed(1)
neural.price.mode1 <- nnet(dji_close ~ Adjunclose, data = CloseL1.train,
                           size = 100, decay = 0.00001, linout = 1, skip=TRUE, MaxNWts=2600, trace=FALSE)
#Predict Close Price on test data
ann_forecast <- forecast(neural.price.model, h= 30, PI = TRUE)
```

Finally, we plot out the line chart to compare them.

```{R}
plot(ann_forecast)
plot(CloseL1)
#calculate RMS error
#Rmse <- sqrt(mean((CloseL1.test-CloseL1$Lag.1)^2))
#Rmse
```



